---
title: "Load and explore the data"
teaching: 15
exercises: 30
questions:
- "What data are required for eqtl mapping?"
objectives:
- To specify which input files are required for qtl2 and how they should be formatted.
keypoints:
- "QTL mapping data consists of a set of tables of data: marker genotypes, phenotypes, marker maps, etc."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```

Load the libraries.
```{r, load_tidyverse}
library(tidyverse)
library(knitr)
library(GGally)
```

## Clinical Phenotypes

The data used in these analyses are available from 
[Data Dryad](https://doi.org/10.5061/dryad.pj105).
From File S1 Attie_eQTL_paper_physiology.Rmd compliments of Daniel Gatti.
Load in the clinical phenotypes.

```{r load_phenos}
# load the data
load("../data/pheno_clin_v6.RData")
```

Subset the phenotypes to include only those analyzed in the paper.

```{r subset_phenos}
# subset the phenotypes to only those addressed in the paper
pheno2keep <- pheno_clin %>% 
  select(weight_14wk, weight_10wk, weight_6wk,
         Ins_14wk, Ins_10wk, Ins_6wk, Ins_per_islet, Ins_iAUC,
         HOMA_IR_0min, HOMA_B_0min,
         Glu_14wk, Glu_10wk, Glu_6wk, Glu_iAUC,
         TG_14wk, TG_10wk, TG_6wk,
         food_ave, WPIC, num_islets)
pheno_clin <- pheno_clin[,c(1:11, 166, 
                            which(colnames(pheno_clin) %in% names(pheno2keep)))]

# convert sex and DO wave (batch) to factors
pheno_clin$sex = factor(pheno_clin$sex)
pheno_clin$DOwave = factor(pheno_clin$DOwave)
```

### QA/QC

#### Proportion Missing Data

```{r pheno_missing_data, warning=FALSE}
tmp = pheno_clin %>% 
  mutate_all(is.na) %>% 
  summarize_all(mean) %>%
  gather(phenotype, value)
kable(tmp, caption = "Proportion of Missing Data")
```

The phenotypes that we're mapping (on the right) are mostly free of missing 
values. The highest are `Ins_per_islet` and WPIC at 3.6%.

#### Phenotype Ranges

```{r pheno_ranges}
tmp = pheno_clin %>%
  select(num_islets:weight_10wk) %>%
  summarize_all(funs(min, max), na.rm = TRUE) %>%
  gather(phenotype, value) %>%
  mutate(phenotype = str_replace(phenotype, "_min", ".min")) %>%
  mutate(phenotype = str_replace(phenotype, "_max", ".max")) %>%
  separate(phenotype, c("phenotype", "stat"), sep = "\\.") %>%
  mutate(stat = factor(stat, levels = c("min", "max"))) %>%
  spread(key = stat, value = value)
kable(tmp, caption = "Phenotype Ranges")
```

#### Univariate Boxplot

```{r pheno_boxplot, warning=FALSE}
pheno_clin %>%
  select(num_islets:weight_10wk) %>%
  gather(phenotype, value) %>%
  ggplot(aes(x = phenotype, y = value)) +
    geom_boxplot() +
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(title = "Distribution of Log Transformed Phenotypes")
```

Log transform and standardize each phenotype. Consider setting points that are 
more than 5 std. dev. from the mean to NA. Only do this if the final 
distribution doesn't look skewed.

```{r pheno_std, warning=FALSE}
pheno_clin_log = pheno_clin %>%
                   mutate_if(is.numeric, log)
pheno_clin_std = pheno_clin_log %>%
                   select(mouse, num_islets:weight_10wk) %>%
                   mutate_if(is.numeric, scale)
pheno_clin_std %>%
  select(num_islets:weight_10wk) %>%
  gather(phenotype, value) %>%
  ggplot(aes(x = phenotype, y = value)) +
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(title = "Distribution of Standardized Phenotypes")
```

```{r outliers, warning=FALSE}
outliers = pheno_clin_std %>% 
              gather(pheno, value, -mouse) %>%
              filter(abs(value) > 5)
kable(outliers, caption = "Potential Outliers")
```

### All Pairs

```{r pheno_all_pairs, warning=FALSE, fig.width=10, fig.height=10}
ggpairs(select(pheno_clin_log, num_islets:weight_10wk)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

The HOMA phenotypes have odd distributions.

```{r homair_vs_homab, warning=FALSE}
saveRDS(pheno_clin_log, 
        file = paste0(data.dir, "pheno_clin_log_outliers_removed.rds"))
ggplot(pheno_clin_log, aes(HOMA_IR_0min, HOMA_B_0min, color = DOwave, shape = sex)) +
  geom_point()
```

There doesn't appear to be a batch effect, but there are a large number of low 
values. Is there some lower bound to the HOMA measurements?

### Figure 1 Boxplots

```{r bw_boxplot, warning=FALSE}
pheno_clin %>%
  select(mouse, sex, starts_with("weight")) %>%
  gather(week, value, -mouse, -sex) %>%
  separate(week, c("tmp", "week")) %>%
  mutate(week = factor(week, levels = c("2wk", "6wk", "10wk"))) %>%
  ggplot(aes(week, value, fill = sex)) +
    geom_boxplot() +
    scale_y_log10() +
    labs(title = "Body Weight", y = "Weight")
```

```{r glucose_boxplot, warning=FALSE}
pheno_clin %>%
  select(mouse, sex, starts_with("Glu")) %>%
  select(mouse, sex, ends_with("wk")) %>%
  gather(week, value, -mouse, -sex) %>%
  separate(week, c("tmp", "week")) %>%
  mutate(week = factor(week, levels = c("6wk", "10wk", "14wk"))) %>%
  ggplot(aes(week, value, fill = sex)) +
    geom_boxplot() +
    scale_y_log10() +
    labs(title = "Glucose", y = "Glucose")
```

```{r insulin_boxplot, warning=FALSE}
pheno_clin %>%
  select(mouse, sex, starts_with("Ins")) %>%
  select(mouse, sex, ends_with("wk")) %>%
  gather(week, value, -mouse, -sex) %>%
  separate(week, c("tmp", "week")) %>%
  mutate(week = factor(week, levels = c("6wk", "10wk", "14wk"))) %>%
  ggplot(aes(week, value, fill = sex)) +
    geom_boxplot() +
    scale_y_log10() +
    labs(title = "Insulin", y = "Insulin")
```

```{r trig_boxplot, warning=FALSE}
pheno_clin %>%
  select(mouse, sex, starts_with("TG")) %>%
  select(mouse, sex, ends_with("wk")) %>%
  gather(week, value, -mouse, -sex) %>%
  separate(week, c("tmp", "week")) %>%
  mutate(week = factor(week, levels = c("6wk", "10wk", "14wk"))) %>%
  ggplot(aes(week, value, fill = sex)) +
    geom_boxplot() +
    scale_y_log10() +
    labs(title = "TG", y = "TG")
```

```{r fig1_boxplots, warning=FALSE}
pheno_clin %>%
  select(mouse, sex, num_islets:Ins_tAUC, food_ave) %>%
  gather(phenotype, value, -mouse, -sex) %>%
  ggplot(aes(sex, value, fill = sex)) +
    geom_boxplot() +
    scale_y_log10() +
    facet_wrap(~phenotype, scales = "free_y")
```

### Tests for sex, wave and diet_days.

```{r sex_diet_wave_anova}
tmp = pheno_clin_log %>%
        select(mouse, sex, DOwave, diet_days, num_islets:weight_10wk) %>%
        gather(phenotype, value, -mouse, -sex, -DOwave, -diet_days) %>%
        group_by(phenotype) %>%
        nest()
mod_fxn = function(df) {
  lm(value ~ sex + DOwave + diet_days, data = df)
}
tmp = tmp %>%
  mutate(model = map(data, mod_fxn)) %>%
  mutate(summ = map(model, tidy)) %>%
  unnest(summ)
kable(tmp, caption = "Effects of Sex, Wave & Diet Days on Phenotypes")
```

```{r sex_diet_wave_effects}
tmp %>%
  filter(term != "(Intercept)") %>%
  mutate(neg.log.p = -log10(p.value)) %>%
  ggplot(aes(term, neg.log.p)) +
    geom_point() +
    facet_wrap(~phenotype) +
    labs(title = "Significance of Sex, Wave & Diet Days on Phenotypes") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
rm(tmp)
```

### Weight vs. Food Intake

```{r bw_vs_food, warning=FALSE}
pheno_clin_log %>%
  select(mouse, sex, food_ave:weight_10wk) %>%
  gather(phenotype, value, -mouse, -sex, -food_ave) %>%
  separate(phenotype, c("phenotype", "week")) %>%
  mutate(week = factor(week, levels = c("2wk", "6wk", "10wk"))) %>%
  ggplot(aes(food_ave, value, color = sex)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(title = "Food Intake vs. Body Weight", y = "log(Body Weight)") + 
    facet_wrap(~week)
```

```{r bw_sex_food_model}
model_fxn = function(df) { lm(value ~ sex*food_ave, data = df) }
tmp = pheno_clin_log %>%
  select(mouse, sex, food_ave:weight_10wk) %>%
  gather(phenotype, value, -mouse, -sex, -food_ave) %>%
  separate(phenotype, c("phenotype", "week")) %>%
  mutate(week = factor(week, levels = c("2wk", "6wk", "10wk"))) %>%
  group_by(week) %>%
  nest() %>%
  mutate(model = map(data, model_fxn)) %>%
  mutate(summ = map(model, tidy)) %>%
  unnest(summ) %>%
  filter(term != "(Intercept)") %>%
  mutate(p.adj = p.adjust(p.value))
kable(tmp, caption = "Effects of Sex and Food Intake on Body Weight")
```

### Correlation Plots

Females

```{r, fig.width=12, fig.height=12, message=FALSE, warning=FALSE}
tmp = pheno_clin_log %>% 
        filter(sex == "F") %>%
        select(num_islets:weight_10wk)
tmp = cor(tmp, use = "pairwise")
pdf(paste0(fig.dir, "phenotype_corr_female.pdf"), width = 12, height = 12)
corrplot.mixed(tmp, upper = "ellipse", lower = "number", 
               main = "Female Clinical Phenotype Correlation")
dev.off()
corrplot.mixed(tmp, upper = "ellipse", lower = "number", 
               main = "Female Clinical Phenotype Correlation")
```

Males

```{r, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}
tmp = pheno_clin_log %>% 
        filter(sex == "M") %>%
        select(num_islets:weight_10wk)
tmp = cor(tmp, use = "pairwise")
pdf(paste0(fig.dir, "phenotype_corr_male.pdf"), width = 12, height = 12)
corrplot.mixed(tmp, upper = "ellipse", lower = "number", 
               main = "Male Clinical Phenotype Correlation")
dev.off()
corrplot.mixed(tmp, upper = "ellipse", lower = "number", 
               main = "Male Clinical Phenotype Correlation")
```

